A tutorial on Bayesian and Frequentist Event History Analyses for psychological time-to-event data

### 2024-09-22 manuscript/manuscript.tex
update manuscript

In experimental psychology, it is still standard practice to analyse response times (RTs), saccade latencies, and fixation durations by calculating average performance across a series of trials. However, differences in mean RT conceal when an experimental effect starts, how long it lasts, how it evolves over increasing waiting time, and whether its onset is time-locked to other events. Such information is useful not only for interpretation of the effects, but also for cognitive psychophysiology and computational model selection (Panis, Schmidt, Wolkersdorfer, \& Schmidt, 2020).

@@

In this paper we focus on a distributional method known as discrete-time event history analysis, a.k.a. hazard analysis, duration analysis, failure-time analysis, survival analysis, and transition analysis.
We first provide a brief overview of hazard analysis to orient the reader to the basic concepts and ideas that we will use throughout the paper. However, this will remain relatively short, and for detailed treatment, see Singer and Willett (2003), Allison (1982), and Allison (2010).

We then provide four different tutorials, each of which is written in R code and publicly available on Github and the Open Science Framework (OSF). The tutorials provide hands-on, concrete examples of key parts of the analytical process, so that others can apply the analyses to their own time-to-event data sets. In Tutorial 1 we illustrate how to calculate the descriptive statistics for a published data set when there is one independent variable. The descriptive statistics are plotted, and we comment on their interpretation. In Tutorial 2 we illustrate how one can fit Bayesian hazard models to the data. We discuss possible link functions, and plot the model-based effects of our predictors of interest. In Tutorial 3 we illustrate how to fit hazard models in a frequentist framework. Even though both frameworks generate similar parameter estimates, we note that model convergence is often not obtained in the frequentist framework. In Tutorial 4 we illustrate how to calculate the descriptive statistics when there are two independent variables.

@@

Similarly, after dividing time in discrete, contiguous time bins indexed by t, let RT be a discrete random variable denoting the rank of the time bin in which a particular person's response occurs in a particular experimental condition. Discrete-time EHA focuses on the discrete-time hazard function h(t) = P(RT = t\textbar{} RT \(\geq\) t) and the discrete-time survivor function S(t) = P(RT \(>\) t) = {[}1-h(t){]}.{[}1-h(t-1){]}.{[}1-h(t-2){]}\ldots{[}1-h(1){]}, and not on the probability mass function p(t) = h(t).S(t-1) and the cumulative distribution function F(t) = 1-S(t). The discrete-time hazard probability function gives you the probability that the event occurs (sometime) in bin t, given that the event has not occurred yet in previous bins. Unlike the discrete-time hazard function, which assesses the unique risk associated with each time bin, the discrete-time survivor function cumulates the bin-by-bin risks of event \emph{non}occurrence.

@@

We focus on factorial within-subject designs in which a large number of observations are made on a relatively small number of participants (small-\emph{N} designs). This approach emphasizes the precision and reproducibility of data patterns at the individual participant level to increase the inferential validity of the design (Baker et al., 2021; Smith \& Little, 2018). In contrast to the large-\emph{N} design that averages across many participants without being able to scrutinize individual data patterns, small-\emph{N} designs retain crucial information about the data patterns of individual observers. This is of great advantage whenever participants differ systematically in their strategies or in the time-courses of their effects, so that blindly averaging them would lead to misleading data patterns. Indeed, Smith and Little (2018) argue that, ``if psychology is to be a mature quantitative science, then its primary theoretical aim should be to investigate systematic functional relationships as they are manifested at the individual participant level'' (p.~2083). Note that because statistical power derives both from the number of participants and from the number of repeated measures per participant and condition, small-\emph{N} designs can have excellent power (Baker et al., 2021; Smith \& Little, 2018).

We used R (Version 4.4.0; R Core Team, 2024)\footnote{We, furthermore, used the R-packages \emph{bayesplot} (Version 1.11.1; Gabry, Simpson, Vehtari, Betancourt, \& Gelman, 2019), \emph{brms} (Version 2.21.0; Bürkner, 2017, 2018, 2021), \emph{citr} (Version 0.3.2; Aust, 2019), \emph{cmdstanr} (Version 0.8.1.9000; Gabry, Češnovar, Johnson, \& Bronder, 2024), \emph{dplyr} (Version 1.1.4; Wickham, François, Henry, Müller, \& Vaughan, 2023), \emph{forcats} (Version 1.0.0; Wickham, 2023a), \emph{ggplot2} (Version 3.5.1; Wickham, 2016), \emph{lme4} (Version 1.1.35.5; Bates, Mächler, Bolker, \& Walker, 2015), \emph{lubridate} (Version 1.9.3; Grolemund \& Wickham, 2011), \emph{Matrix} (Version 1.7.0; Bates, Maechler, \& Jagan, 2024), \emph{nlme} (Version 3.1.166; Pinheiro \& Bates, 2000), \emph{papaja} (Version 0.1.2.9000; Aust \& Barth, 2023), \emph{patchwork} (Version 1.2.0; Pedersen, 2024), \emph{purrr} (Version 1.0.2; Wickham \& Henry, 2023), \emph{RColorBrewer} (Version 1.1.3; Neuwirth, 2022), \emph{Rcpp} (Eddelbuettel \& Balamuta, 2018; Version 1.0.12; Eddelbuettel \& François, 2011), \emph{readr} (Version 2.1.5; Wickham, Hester, \& Bryan, 2024), \emph{RJ-2021-048} (Bengtsson, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021), \emph{standist} (Version 0.0.0.9000; Girard, 2024), \emph{stringr} (Version 1.5.1; Wickham, 2023b), \emph{tibble} (Version 3.2.1; Müller \& Wickham, 2023), \emph{tidybayes} (Version 3.0.6; Kay, 2023), \emph{tidyr} (Version 1.3.1; Wickham, Vaughan, \& Girlich, 2024), \emph{tidyverse} (Version 2.0.0; Wickham et al., 2019), and \emph{tinylabels} (Version 0.2.4; Barth, 2023).} for all reported analyses. Web links are printed in bold. The content of the tutorials is mainly based on Allison (2010), Singer and Willett (2003), McElreath (2018), Kurz (2023a), and Kurz (2023b).

@@

The symmetric logit link function transforms a (hazard) probability into the log of the odds ratio. The asymmetric complementary log-log (cloglog) link function transforms hazard into the logarithm of the negated logarithm of the probability of event \emph{non}occurrence. An important difference between these two link functions is that cloglog provides a discrete-time hazard model that has a built-in proportional hazards assumption, while logit provides a proportional odds assumption (see below). The cloglog link is preferred over the logit link when events can occur in principle at any time point within a bin, which is the case for RT data (Singer \& Willett, 2003). Third, one has to choose a specification of the effect of discrete TIME (i.e., the time bin index t). One can choose a general specification (one intercept per bin) or a functional specification, such as a polynomial one (compare model 1 with models 2, 3, and 4 below).

An example (single-level) discrete-time hazard model with three predictors (TIME, X1, X2) and the cloglog link function can be written as follows:

@@

In the case of a large-\emph{N} design without repeated measurements, the parameters of a discrete-time hazard model can be estimated using standard logistic regression software (after expanding the typical person-trial-oriented data set into a person-trial-bin-oriented data set; Allison (2010)). When there is clustering in the data, as in the case of a small-\emph{N} design with repeated measurements, the parameters of a discrete-time hazard model can be estimated using population-averaged methods (e.g., Generalized Estimating Equations), and Bayesian or frequentist generalized linear mixed models (Allison, 2010).

@@

To gain a sense of what prior \emph{logit} values would approximate a uniform distribution on the probability scale, Kurz (2023a) simulated a large number of draws from the Uniform(0,1) distribution, converted those draws to the log-odds metric, and fitted a Student's t distribution. Row C in Figure 4 shows that using a t-distribution with 7.61 degrees of freedom and a scale parameter of 1.57 as a prior on the logit scale, approximates a uniform distribution on the probability scale. According to Kurz (2023a), such a prior might be a good prior for the intercept(s) in a logit-hazard model, while the N(0,1) prior in row D might be a good prior for the beta parameters in a logit-hazard model, as it gently regularizes p towards .5 (i.e., a zero effect on the logit scale).

To gain a sense of what prior \emph{cloglog} values would approximate a uniform distribution on the hazard probability scale, we followed Kurz's approach and simulated a large number of draws from the Uniform(0,1) distribution, converted them to the cloglog metric, and fitted a skew-normal model (due to the asymmetry of the cloglog link function). Row E shows that using a skew-normal distribution with a mean of -0.59, a standard deviation of 1.26, and a skewness of -4.22 as a prior on the cloglog scale, approximates a uniform distribution on the probability scale.
However, because hazard values below .5 are more likely in RT studies, using a skew-normal distribution with a mean of -1, a standard deviation of 1, and a skewness of -2 as a prior on the cloglog scale (row F), might be a good weakly informative prior for the intercept(s) in a cloglog-hazard model.
A skew-normal distribution with a mean of -0.2, a standard deviation of 0.71, and a skewness of -2.2 might be a good weakly informative prior for the beta parameters in a cloglog-hazard model as it gently regularizes p towards .6321 (i.e., a zero effect on the cloglog scale).

@@

Based on previous work (Panis, 2020; Panis, Moran, et al., 2020; Panis \& Schmidt, 2022; Panis et al., 2017; Panis \& Wagemans, 2009) and because cognition is likely the behavior of a non-linear dynamical system {[}ref{]}, we relax all three assumptions in model 4. We use the same data set and priors as for model 2.

@@

In Figure 7 we compere the parameter estimates of model M3 from brms() with those of glmer().

@@

\caption{Parameter estimates for model M3 from brms() and glmer().}\label{fig:plot-comparison}
\end{figure}

@@

\caption{Estimated discrete-time hazard, survivor, and conditional accuracy functions for participant 6, as a function of prime type (blank, congruent, incongruent) and mask type (no mask, relevant, irrelevant, lines).}\label{fig:plot-2ivs}
\end{figure}

@@

We noticed that many researchers are still reluctant to abandon analysis-of-variance and switch to event history analysis when analyzing time-to-event data. By providing this tutorial, we hope that researchers will start using hazard analysis more often, due to the many advantages described below. While we focus here on within-subject, factorial, small-\emph{N} designs, it is important to realize that event history analysis can be applied to other designs as well (large-\emph{N} designs with only one measurement per subject, between-subject designs, etc.).

@@

Statisticians and mathematical psychologists recommend focusing on the hazard function when analyzing time-to-event data for various reasons. First, as discussed by Holden, Van Orden, and Turvey (2009), ``probability density functions can appear nearly identical, both statistically and to the naked eye, and yet are clearly different on the basis of their hazard functions (but not vice versa). Hazard functions are thus more diagnostic than density functions'' (p.~331) when one is interested in studying the detailed shape of a RT distribution (see also Figure 1 in \ldots.).

Second, because RT distributions may differ from one another in multiple ways, Townsend (1990) developed a dominance hierarchy of statistical differences between two arbitrary distributions A and B. For example, if F\textsubscript{A}(t) \textgreater{} F\textsubscript{B}(t) for all t, then both cumulative distribution functions are said to show a complete ordering. Townsend (1990) showed that a complete ordering on the hazard functions ---\(\lambda\)\textsubscript{A}(t) \textgreater{} \(\lambda\)\textsubscript{B}(t) for all t--- implies a complete ordering on both the cumulative distribution and survivor functions ---F\textsubscript{A}(t) \textgreater{} F\textsubscript{B}(t) and S\textsubscript{A}(t) \textless{} S\textsubscript{B}(t)--- which in turn implies an ordering on the mean latencies ---mean A \textless{} mean B. In contrast, an ordering on two means does \emph{not} imply a complete ordering on the corresponding F(t) and S(t) functions, and a complete ordering on these latter functions does \emph{not} imply a complete ordering on the corresponding hazard functions. This means that stronger conclusions can be drawn from data when comparing the hazard functions using EHA. For example, when mean A \textless{} mean B, the hazard functions might show a complete ordering (i.e., for all t), a partial ordering (e.g., only for t \textgreater{} 300 ms, or only for t \textless{} 500 ms), or they may cross each other one or more times.
Thus, because the discrete-time hazard function identifies unique information about event occurrence in each bin, while F(t) accumulates the complements of the hazard estimates from the current and previous bins -- S(t) = P(RT \(>\) t) = {[}1-h(t){]}.{[}1-h(t-1){]}.{[}1-h(t-2){]}\ldots{[}1-h(1){]} --, instead of using delta-plots -- differences in quantiles from F(t)\(^-1\) -- one can simply plot delta-h(t) functions (see ref).

@@

Finally, as explained by Kelso, Dumas, and Tognoli (2013), it is crucial to first have a precise description of the macroscopic behavior of a system (here: h(t) and ca(t) functions) in order to know what to derive on the microscopic level. EHA can thus solve the problem of model mimicry, i.e., the fact that different computational models predict the same mean RTs as observed in the empirical data, but not necessarily the detailed shapes of the empirical RT hazard distributions. Also, fitting parametric functions or computational models to data without studying the shape of the empirical discrete-time h(t) and ca(t) functions can miss important features in the data (Panis, Moran, et al., 2020; Panis \& Schmidt, 2016).

@@

One important issue is that of possible individual differences in the overall location of the distribution, and the time course of psychological effects. For example, when you wait for a response of the participant on each trial, you allow the participant to have control over the trial duration, and some participants might respond only when they are confident that their emitted response will be correct. These issues can be avoided by introducing a (relatively short) response deadline in each trial, e.g., 600 ms for simple detection tasks, 1000 ms for more difficult discrimination tasks, or 2 s for tasks requiring extended high-level processing. Because EHA can deal in a straightforward fashion with right-censored observations (i.e., trials without an observed response), introducing a response deadline is recommended when designing RT experiments. Furthermore, introducing a response deadline and asking participants to respond before the deadline as much as possible, will also lead to individual distributions that overlap in time, which is important when selecting a common analysis time window when fitting hazard models.
But even when using a response deadline, participants can differ qualitatively in the effects they display (refs). One way to deal with this is to describe and interpret the different patterns. Another way is to run a clustering algorithm on the individual hazard estimates across all conditions. The obtained dendrogram can then be used to identify (hopefully big) clusters of participants that behave similarly, and to identify (hopefully small) clusters of participants with outlying behavioral patterns. One might then exclude the outlying participants before fitting a hazard model.

\subsection{Limitations}\label{limitations}

@@

The hazard models in this tutorial assume that there is one event of interest. For RT data, this event consitutes a single transition between an ``idle'' state and a ``responded'' state. However, in certain situations, more than one event of interest might exist. For example, an individual might transition back and forth between a ``healthy'' state and a ``depression'' state, before being absorbed in a final ``death'' state. When you have data on the timing of these transitions, one can apply multi-state models which generalize survival analysis to transitions between three or more states (refs).
Also, the predictor variables in this tutorial are time-invariant, i.e., their value did not change over the course of a trial. Another extension is to include time-varying predictors, i.e., predictors whose value can change across the time bins within a trial.

@@

see paper\ldots{}

@@

>>>>

In experimental psychology, it is still standard practice to analyse response times (RTs), saccade latencies, and fixation durations by calculating average performance across a series of trials. However, differences in mean RT conceal when an experimental effect starts, how long it lasts, how it evolves with increasing waiting time, and whether its onset is time-locked to other events. Such information is useful not only for interpretation of the effects, but also for cognitive psychophysiology and computational model selection (Panis, Schmidt, Wolkersdorfer, \& Schmidt, 2020).

@@

In this paper we focus on a distributional method known as discrete-time event history analysis, a.k.a. hazard analysis, duration analysis, failure-time analysis, survival analysis, and transition analysis.
We first provide a brief overview of hazard analysis to orient the reader to the basic concepts that we will use throughout the paper. However, this will remain relatively short, and for detailed treatment, see Singer and Willett (2003), Allison (1982), and Allison (2010).

We then provide four different tutorials, each of which is written in R code and publicly available on Github and the Open Science Framework (OSF). The tutorials provide hands-on, concrete examples of key parts of the analytical process, so that others can apply the analyses to their own time-to-event data sets. In Tutorial 1 we illustrate how to calculate the descriptive statistics for a published RT data set when there is one independent variable. The descriptive statistics are plotted, and we comment on their interpretation. In Tutorial 2 we illustrate how one can fit Bayesian hazard models to the data. We discuss possible link functions, and plot the model-based effects of our predictors of interest. In Tutorial 3 we illustrate how to fit hazard models in a frequentist framework. Even though both frameworks generate similar parameter estimates, we note that model convergence is often not obtained in the frequentist framework. In Tutorial 4 we illustrate how to calculate the descriptive statistics when there are two independent variables.

@@

Similarly, after dividing time in discrete, contiguous time bins indexed by t, let RT be a discrete random variable denoting the rank of the time bin in which a particular person's response occurs in a particular experimental condition. Discrete-time EHA focuses on the discrete-time hazard function h(t) = P(RT = t\textbar{} RT \(\geq\) t) and the discrete-time survivor function S(t) = P(RT \(>\) t) = {[}1-h(t){]}.{[}1-h(t-1){]}.{[}1-h(t-2){]}\ldots{[}1-h(1){]}, and not on the probability mass function P(t) = P(RT = t) = h(t).S(t-1) nor the cumulative distribution function F(t) = 1-S(t). The discrete-time hazard probability gives you the probability that the event occurs (sometime) in bin t, given that the event has not occurred yet in previous bins. Unlike the discrete-time hazard function, which assesses the unique risk of event occurrencce associated with each time bin, the discrete-time survivor function cumulates the bin-by-bin risks of event \emph{non}occurrence to obtain the probability that the event occurs afer bin t.

@@

Here we focus on factorial within-subject designs in which a large number of observations are made on a relatively small number of participants (small-\emph{N} designs). This approach emphasizes the precision and reproducibility of data patterns at the individual participant level to increase the inferential validity of the design (Baker et al., 2021; Smith \& Little, 2018). In contrast to the large-\emph{N} design that averages across many participants without being able to scrutinize individual data patterns, small-\emph{N} designs retain crucial information about the data patterns of individual observers. This is of great advantage whenever participants differ systematically in their strategies or in the time-courses of their effects, so that blindly averaging them would lead to misleading data patterns. Indeed, Smith and Little (2018) argue that, ``if psychology is to be a mature quantitative science, then its primary theoretical aim should be to investigate systematic functional relationships as they are manifested at the individual participant level'' (p.~2083). Note that because statistical power derives both from the number of participants and from the number of repeated measures per participant and condition, small-\emph{N} designs can have excellent power (Baker et al., 2021; Smith \& Little, 2018).

We used R (Version 4.4.0; R Core Team, 2024)\footnote{We, furthermore, used the R-packages \emph{bayesplot} (Version 1.11.1; Gabry, Simpson, Vehtari, Betancourt, \& Gelman, 2019), \emph{brms} (Version 2.21.0; Bürkner, 2017, 2018, 2021), \emph{citr} (Version 0.3.2; Aust, 2019), \emph{cmdstanr} (Version 0.8.1.9000; Gabry, Češnovar, Johnson, \& Bronder, 2024), \emph{dplyr} (Version 1.1.4; Wickham, François, Henry, Müller, \& Vaughan, 2023), \emph{forcats} (Version 1.0.0; Wickham, 2023a), \emph{ggplot2} (Version 3.5.1; Wickham, 2016), \emph{lme4} (Version 1.1.35.5; Bates, Mächler, Bolker, \& Walker, 2015), \emph{lubridate} (Version 1.9.3; Grolemund \& Wickham, 2011), \emph{Matrix} (Version 1.7.0; Bates, Maechler, \& Jagan, 2024), \emph{nlme} (Version 3.1.166; Pinheiro \& Bates, 2000), \emph{papaja} (Version 0.1.2.9000; Aust \& Barth, 2023), \emph{patchwork} (Version 1.2.0; Pedersen, 2024), \emph{purrr} (Version 1.0.2; Wickham \& Henry, 2023), \emph{RColorBrewer} (Version 1.1.3; Neuwirth, 2022), \emph{Rcpp} (Eddelbuettel \& Balamuta, 2018; Version 1.0.12; Eddelbuettel \& François, 2011), \emph{readr} (Version 2.1.5; Wickham, Hester, \& Bryan, 2024), \emph{RJ-2021-048} (Bengtsson, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021), \emph{standist} (Version 0.0.0.9000; Girard, 2024), \emph{stringr} (Version 1.5.1; Wickham, 2023b), \emph{tibble} (Version 3.2.1; Müller \& Wickham, 2023), \emph{tidybayes} (Version 3.0.6; Kay, 2023), \emph{tidyr} (Version 1.3.1; Wickham, Vaughan, \& Girlich, 2024), \emph{tidyverse} (Version 2.0.0; Wickham et al., 2019), and \emph{tinylabels} (Version 0.2.4; Barth, 2023).} for all reported analyses. Web links are printed in bold. The content of the tutorials is mainly based on Allison (2010), Singer and Willett (2003), McElreath (2018), Kurz (2023a), and Kurz (2023b).

@@

The symmetric logit link function transforms a (hazard) probability into the log of the odds ratio. The asymmetric complementary log-log (cloglog) link function transforms hazard into the logarithm of the negated logarithm of the probability of event \emph{non}occurrence. An important difference between these two link functions is that cloglog provides a discrete-time hazard model that has a built-in proportional hazards assumption, while logit provides a proportional odds assumption. The cloglog link is preferred over the logit link when events can occur in principle at any time point within a bin, which is the case for RT data (Singer \& Willett, 2003). Third, one has to choose a specification of the effect of discrete TIME (i.e., the time bin index t). One can choose a general specification (one intercept per bin) or a functional specification, such as a polynomial one (compare model 1 with models 2, 3, and 4 below).

An example (single-level) discrete-time hazard model with three predictors (TIME, X\textsubscript{1}, X\textsubscript{2}), the cloglog link function, and a third-order polynomial specification for TIME can be written as follows:

@@

An example (single-level) discrete-time hazard model with a general specification for TIME (separate intercepts for each of six bins, where D1 to D6 are binary variables identifying each bin) and a single predictor (X\textsubscript{1}) can be written as follows:

cloglog{[}h(t){]} = ln(-ln{[}1-h(t){]}) = {[}\(\alpha\)\textsubscript{1}D1 + \(\alpha\)\textsubscript{2}D2 + \(\alpha\)\textsubscript{3}D3 + \(\alpha\)\textsubscript{4}D4 + \(\alpha\)\textsubscript{5}D5 + \(\alpha\)\textsubscript{6}D6{]} + {[}\(\beta\)\textsubscript{1}X\textsubscript{1}{]}.

In the case of a large-\emph{N} design without repeated measurements, the parameters of a discrete-time hazard model can be estimated using standard logistic regression software after expanding the typical person-trial-oriented data set into a person-trial-bin-oriented data set (Allison, 2010). When there is clustering in the data, as in the case of a small-\emph{N} design with repeated measurements, the parameters of a discrete-time hazard model can be estimated using population-averaged methods (e.g., Generalized Estimating Equations), and Bayesian or frequentist generalized linear mixed models (Allison, 2010).

@@

To gain a sense of what prior \emph{logit} values would approximate a uniform distribution on the probability scale, Kurz (2023a) simulated a large number of draws from the Uniform(0,1) distribution, converted those draws to the log-odds metric, and fitted a Student's t distribution. Row C in Figure 4 shows that using a t-distribution with 7.61 degrees of freedom and a scale parameter of 1.57 as a prior on the logit scale, approximates a uniform distribution on the probability scale. According to Kurz (2023a), such a prior might be a good prior for the intercept(s) in a logit-hazard model, while the N(0,1) prior in row D might be a good prior for the non-intercept parameters in a logit-hazard model, as it gently regularizes p towards .5 (i.e., a zero effect on the logit scale).

To gain a sense of what prior \emph{cloglog} values would approximate a uniform distribution on the hazard probability scale, we followed Kurz's approach and simulated a large number of draws from the Uniform(0,1) distribution, converted them to the cloglog metric, and fitted a skew-normal model (due to the asymmetry of the cloglog link function). Row E shows that using a skew-normal distribution with a mean of -0.59, a standard deviation of 1.26, and a skewness of -4.22 as a prior on the cloglog scale, approximates a uniform distribution on the probability scale.
However, because hazard values below .5 are more likely in RT studies, using a skew-normal distribution with a mean of -1, a standard deviation of 1, and a skewness of -2 as a prior on the cloglog scale (row F), might be a good weakly informative prior for the intercept(s) in a cloglog-hazard model.
A skew-normal distribution with a mean of -0.2, a standard deviation of 0.71, and a skewness of -2.2 might be a good weakly informative prior for the non-intercept parameters in a cloglog-hazard model as it gently regularizes p towards .6321 (i.e., a zero effect on the cloglog scale).

@@

Based on previous work (Panis, 2020; Panis, Moran, et al., 2020; Panis \& Schmidt, 2022; Panis et al., 2017; Panis \& Wagemans, 2009), we relax all three assumptions in model 4. We use the same data set and priors as for model 2.

@@

In Figure 7 we compere the parameter estimates of model M3 from brm() with those of glmer().

@@

\caption{Parameter estimates for model M3 from brm() and glmer().}\label{fig:plot-comparison}
\end{figure}

@@

\caption{Sample-based discrete-time hazard, survivor, and conditional accuracy functions for participant 6, as a function of prime type (blank, congruent, incongruent) and mask type (no mask, relevant, irrelevant, lines).}\label{fig:plot-2ivs}
\end{figure}

@@

We noticed that many researchers are still reluctant to abandon analysis-of-variance and switch to event history analysis when analyzing time-to-event data. By providing this tutorial, we hope that researchers will start using descriptive and inferential hazard analysis more often, due to the many advantages described below. While we focused here on within-subject, factorial, small-\emph{N} designs, it is important to realize that event history analysis can be applied to other designs as well (large-\emph{N} designs with only one measurement per subject, between-subject designs, etc.). We also discuss the issue of individual differences, limitations, and extensions.

@@

Statisticians and mathematical psychologists recommend focusing on the hazard function when analyzing time-to-event data for various reasons. First, as discussed by Holden, Van Orden, and Turvey (2009), ``probability density functions can appear nearly identical, both statistically and to the naked eye, and yet are clearly different on the basis of their hazard functions (but not vice versa). Hazard functions are thus more diagnostic than density functions'' (p.~331) when one is interested in studying the detailed shape of a RT distribution (see also Figure 1 in Panis, Schmidt, et al., 2020).

Second, because RT distributions may differ from one another in multiple ways, Townsend (1990) developed a dominance hierarchy of statistical differences between two arbitrary distributions A and B. For example, if F\textsubscript{A}(t) \textgreater{} F\textsubscript{B}(t) for all t, then both cumulative distribution functions are said to show a complete ordering. Townsend (1990) showed that a complete ordering on the hazard functions ---\(\lambda\)\textsubscript{A}(t) \textgreater{} \(\lambda\)\textsubscript{B}(t) for all t--- implies a complete ordering on both the cumulative distribution and survivor functions ---F\textsubscript{A}(t) \textgreater{} F\textsubscript{B}(t) and S\textsubscript{A}(t) \textless{} S\textsubscript{B}(t)--- which in turn implies an ordering on the mean latencies ---mean A \textless{} mean B. In contrast, an ordering on two means does \emph{not} imply a complete ordering on the corresponding F(t) and S(t) functions, and a complete ordering on these latter functions does \emph{not} imply a complete ordering on the corresponding hazard functions. This means that stronger conclusions can be drawn from data when comparing the hazard functions using EHA. For example, when mean A \textless{} mean B, the hazard functions might show a complete ordering (i.e., for all t), a partial ordering (e.g., only for t \textgreater{} 300 ms, or only for t \textless{} 500 ms), or they may cross each other one or more times.
As a result, instead of using delta-plots for RT -- differences in quantiles from F(t)\(^-1\) -- one can simply plot delta-h(t) functions (see Panis, 2020).

@@

Finally, as explained by Kelso, Dumas, and Tognoli (2013), it is crucial to first have a precise description of the macroscopic behavior of a system (here: h(t) and ca(t) functions) in order to know what to derive on the microscopic level. EHA can thus solve the problem of model mimicry, i.e., the fact that different computational models can often predict the same mean RTs as observed in the empirical data, but not necessarily the detailed shapes of the empirical RT hazard distributions. Also, fitting parametric functions or computational models to data without studying the shape of the empirical discrete-time h(t) and ca(t) functions can miss important features in the data (Panis, Moran, et al., 2020; Panis \& Schmidt, 2016).

@@

One important issue is that of possible individual differences in the overall location of the distribution, and the time course of psychological effects. For example, when you wait for a response of the participant on each trial, you allow the participant to have control over the trial duration, and some participants might respond only when they are confident that their emitted response will be correct. These issues can be avoided by introducing a (relatively short) response deadline in each trial, e.g., 600 ms for simple detection tasks, 1000 ms for more difficult discrimination tasks, or 2 s for tasks requiring extended high-level processing. Because EHA can deal in a straightforward fashion with right-censored observations (i.e., trials without an observed response), introducing a response deadline is recommended when designing RT experiments. Furthermore, introducing a response deadline and asking participants to respond before the deadline as much as possible, will also lead to individual distributions that overlap in time, which is important when selecting a common analysis time window when fitting hazard models.
But even when using a response deadline, participants can differ qualitatively in the effects they display (see Panis, 2020). One way to deal with this is to describe and interpret the different patterns. Another way is to run a clustering algorithm on the individual hazard estimates across all conditions. The obtained dendrogram can then be used to identify a (hopefully big) cluster of participants that behave similarly, and to identify a (hopefully small) cluster of participants with outlying behavioral patterns. One might then exclude the outlying participants before fitting a hazard model.

\subsection{Limitation(s)}\label{limitations}

@@

The hazard models in this tutorial assume that there is one event of interest. For RT data, this event consitutes a single transition between an ``idle'' state and a ``responded'' state. However, in certain situations, more than one event of interest might exist. For example, an individual might transition back and forth between a ``healthy'' state and a ``depression'' state, before being absorbed in a final ``death'' state. When you have data on the timing of these transitions, one can apply multi-state models which generalize survival analysis to transitions between three or more states (Steele, Goldstein, \& Browne, 2004).
Also, the predictor variables in this tutorial are time-invariant, i.e., their value did not change over the course of a trial. Thus, another extension is to include time-varying predictors, i.e., predictors whose value can change across the time bins within a trial.

@@

RT and accuracy distributions are a rich source of information on the time course of cognitive processing. The changing effects of our experimental manipulations with increases in waiting time become strikingly clear when looking at response hazards and microlevel speed-accuracy trade-off functions. We hope that experimental psychologists and cognitive neuroscientists are less reluctant to embrace event history analysis after reading this tutorial.

@@

\bibitem[\citeproctext]{ref-SteeleF2004}
Steele, F., Goldstein, H., \& Browne, W. (2004). A general multilevel multistate competing risks model for event history data, with an application to a study of contraceptive use dynamics. \emph{Statistical Modelling}, \emph{4}(2), 145--159. \url{https://doi.org/10.1191/1471082X04st069oa}
